---
title: "Example: step 1.1 - creating a density decay map"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Example: step 1.1 - creating a density decay map}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: references.bib
csl: apa.csl
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Creating density-decay maps

In cases where there is no local GPS data available users are recommended to estimate a bird density based on a distance-decay function, and use this as their bird distribution map in the model.

In the previous version of SeabORD, which ran on MatLab, the parameters to be used for creating your density-decay map would be specified in the interface. This is no longer incorporated within the code and users need to produce their own map.

We provide an example here of how this can be done, using similar parameters to those within the old version of SeabORD and the Cumulative Effects Framework (CEF). We also demonstrate how to assess the potential displacement that will occur given this map and a set of wind farms so that users can gain a better understanding of potential interaction in a given scenario.

This example is for kittiwakes at the Whinnyfold colony near Aberdeen, and the wind farms used are Aberdeen Bay, Hywind and Kincardine, but it should be relatively straightforward to implement this code with your own colony and wind farm footprints, as all you will need to supply is the colony coordinates and your own set of shapefiles for the wind farms.

**Please note** that the parameters we have used in the distance decay map creation (foraging range, "fr", and percentage in foraging range, "pinfr") are for example sake and you will need to make your own judgments here based on existing knowledge of the species and the locality.

This example will include the following steps:

1.  Housekeeping - loading packages, sourcing functions, and loading data into environment
2.  Create a distance by sea map - required for producing distance decay map
3.  Create distance decay map
4.  Plot the distance decay map with OWFs
5.  Assess the interaction with OWFs

## 1. Housekeeping

Load the required packages:

```{r setup, warning = FALSE, message = FALSE}
packages <- c("ggplot2", "dplyr", "sf", "terra", "gdistance", "ggspatial")
lapply(packages, library, character.only = TRUE)

```

Source the functions:

```{r}
# set the path to where the data and function files are found:
mypath <- "C:/Users/chrpol/Documents/seabord_pkg/seabORD_pkg/densitydecay"
source(file.path(mypath, "functions_densitydecay.R")) 

# now we can see the assorted range of functions uploaded into our environment 
# e.g. "calc_dist_restricted" which is used to calculate the distance to sea cells either with or without wind farms, allowing us to calculate obstructed distance when OWFs are included. 
calc_dist_restricted
```

Load the data into environment:

```{r}

# Load in the sea mask of the UK (shows which cells are on land and sea)
seamask <- terra::rast(file.path(mypath,"seamask_3035.tif"))
#terra::plot(seamask)
searast <- raster::raster(seamask)

# Load in the UK coastline for plotting later:
coastline <- readRDS(file.path(mypath,"cef_coast_3035.rds"))
UK_coastline <- dplyr::filter(coastline, COUNTRY == "United Kingdom")
UK_coastline_LAEA <- st_transform(UK_coastline, crs= "+proj=laea +lat_0=52 +lon_0=10 +x_0=4321000 +y_0=3210000 +ellps=GRS80 +units=m +no_defs" )

# Load in the wind farm polygons. In this case it is a shapefile containing the three wind farms closest to Whinnyfold (Hywind, Aberdeen Bay, Kincardine). You could upload any shapefiles you want - just ensure you transform them to the same coordinate reference system (EPSG:3035) and they are formatted in a similar fashion to the example.
ORDpolys <- st_read(file.path(mypath,"owfpolys.shp"))


# Load the flight gird correction layer as geographic correction is necessary for gdistance calculations of the class Transition as they are made with >4 directions in our instance. This correction layer has been pre-calculated and covers the full map we are using. 
load(file.path(mypath, "FlightGridcorrection_3035.RData")) 


```

## 2. Distance by sea map

First we set the colony location, in this case Whinnyfold, but you could change this to any colony around the UK that you wish to model. **PLEASE NOTE:** these calculations will not work if the set location is on land. It needs to be placed beside or on the sea cells so it is not completely obstructed by land cells which are defined as impassable. We recommend checking this with a plot.

```{r}
# Set the location of our colony at Whinnyfold and transform it to LAEA projection:
colony <- st_as_sf(data.frame(lon = -1.85849, lat = 57.388807), coords = c("lon", "lat"), crs = 4326)
colonyLAEA <- st_transform(colony, crs = "EPSG:3035")  # Convert to LAEA Europe

# check the colony location is beside sea cells:
expand_km <- 2 # how far to expand the plot in km on either side of the colony
ggplot() +
  geom_sf(data=UK_coastline_LAEA, fill = "grey", colour="black") +  # coastline
  geom_sf(data=colonyLAEA, colour="blue", pch=4, size =4)+          # colony
  coord_sf(                                                         # 
    xlim=c(st_coordinates(colonyLAEA)[,1] - (expand_km*1000), st_coordinates(colonyLAEA)[,1] + (expand_km*1000)),
    ylim=c(st_coordinates(colonyLAEA)[,2] - (expand_km*1000), st_coordinates(colonyLAEA)[,2] + (expand_km*1000)), expand = FALSE) +
  annotation_scale(location = "tr", width_hint = 0.5)
  

# Create you distance by sea map using the function calc_dist_restricted
DBS_map <- calc_dist_restricted(mymap=searast,
                                  obspolys=NULL,               # not including ORDs this time
                                  targetcoords = colonyLAEA,   #
                                  FlightGridcorrection_3035 = FlightGridcorrection_3035,
                                  directions=16, obspenalty=1e+08,
                                  maxdist=300)
```

## 3. Distance decay map

## 4. Plotting

## 5. Interaction assessment
