---
title: "Example kittiwake:  step 2.2 - seabORD calibration - GPS maps"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Example kittiwake:  step 2.2 - seabORD calibration - GPS maps}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

Load the seabORD package

```{r setup, warning = FALSE, message = FALSE}

# load seabORD
library(seabORD)
```

```{r, echo = FALSE}
library(magrittr) #used for ggplotting but users won't need this
```

```{r, echo=FALSE}

### UK outline

#uk_map <- dplyr::select(
  #rnaturalearth::ne_countries(country = "United Kingdom",
                              #scale = "medium",
                              #returnclass = "sf"),
  #name,
  #geometry)

#use coastline
uk_map <- seabORD::cef_coast_4326


### theme definition
sysfonts::font_add_google("Montserrat", "montserrat")  # Base and heading font
sysfonts::font_add_google("JetBrains Mono", "jetbrains_mono")  # Code font
showtext::showtext_auto()

theme_bslib <- ggplot2::theme(
  plot.background = ggplot2::element_rect(fill = "#fff", color = NA),       # Background color
  panel.background = ggplot2::element_rect(fill = "#EAEFEC", color = NA),  # Panel background (secondary)
  panel.grid.major = ggplot2::element_line(color = "#EAEFEC"),             # Major grid lines
  panel.grid.minor = ggplot2::element_line(color = "#EAEFEC"),             # Minor grid lines
  axis.text = ggplot2::element_text(color = "#292C2F", family = "montserrat"),  # Axis text (foreground and font)
  axis.title = ggplot2::element_text(color = "#292C2F", family = "montserrat"), # Axis titles
  plot.title = ggplot2::element_text(color = "#0483A4", family = "montserrat", size = 16, face = "bold", hjust = 0.5), # Title
  legend.background = ggplot2::element_rect(fill = "#fff", color = NA),    # Legend background
  legend.key = ggplot2::element_rect(fill = "#EAEFEC", color = NA),        # Legend key
  legend.text = ggplot2::element_text(color = "#292C2F", family = "montserrat"), # Legend text
  plot.caption = ggplot2::element_text(color = "#292C2F", family = "jetbrains_mono"),
  axis.title.x = ggplot2::element_blank(),  # Remove x-axis label
  axis.title.y = ggplot2::element_blank(),# Caption for code font
)
```

# Introduction & background

A key component of the seabORD model is the need to calibrate the prey level inputs used for each new species and colony combination. This currently involves re-running the model without ORDs (baseline simulations) with a range of prey levels to ascertain the range of values within which the model returns adult mass and chick mortality rates consistent with pre-defined values representing “moderate” conditions (Figure 1). This is to allow simulation of the range of outcomes expected under these conditions, as the model is highly sensitive to the prey density parameter.

![*Figure 1: : Schematic plot of kittiwake adult mass loss (A) and chicks per nest (B) withdrawn from 300 baseline simulations (no ORD footprints) ranging from a prey value of 50 up to 250. The red lines are for smoothed model outputs, which are overlaid on grey points indicating the individual estimates. For calibration we have upper and lower bound for each output (A & B) which are picked to reflect “moderate” conditions, i.e., what is observed in a typical year for kittiwakes at the Isle of May, and these are indicates by the dashed grey horizontal lines. The grey shaded blocks indicate where the respective outputs fall within these bound, and the light blue shaded block indicates the prey values where there is overlap between the estimates, which is the prey range used in subsequent simulations with ORDs included.*](images/Calibration_plot_schematic_extended_seabordpkg.png)

## This example

In this example we will run 20 baseline runs (no wind farm footprints included) with a range of prey values and plot the outputs in attempt to capture "moderate" conditions for kittiwakes at the Isle of May.

# Prepare for the run

The inputs for

*[include table here when updated in the scenario example]*

# Calibrate the model

# Plot the outputs 

If the time constraints do not allow for running of a full set of calibration runs within the workshop here is a dataset from calibration runs under the same setting conducted prior to the workshop:

```{r}
# load in the datastet "example_calibration_output":
seabORD::example_calibration_output

#load("C:/Users/chrpol/Documents/seabord_pkg/seabORD_pkg/data/example_calibration_output.rda")

example_calibration_output$output_a0$data <- example_calibration_output$output_a0$data %>%
  mutate(percentage_loss = ((BM_adult_t0.mn - BM_adult.mn)/BM_adult_t0.mn))

massupper <- 0.11
masslower <- 0.09
produpper <- 0.55
prodlower <- 0.45

lo <- 173
hi <- 174

  pAd <- ggplot() +
    #geom_hline(yintercept = 0.1, lty=2) +
    geom_hline(yintercept = massupper, lty=2, colour="darkgrey") +
    geom_hline(yintercept = masslower, lty=2, colour="darkgrey") +
    geom_vline(xintercept = hi, lty=1, colour="black") +
    geom_vline(xintercept = lo, lty=1, colour="black") +
    geom_point(dat = example_calibration_output$output_a0$data,aes(x=Prey, y = percentage_loss) , colour="red") +
    geom_line(dat = example_calibration_output$output_a0$data,aes(x=Prey, y = percentage_loss), colour="red") +
    geom_smooth(dat = example_calibration_output$output_a0$data,aes(x=Prey, y = percentage_loss), colour="orange",method = "loess") +
    labs(y = "Adults % loss")
  
  pCh <- ggplot() +
    #geom_hline(yintercept = 0.5, lty= 2) + 
    geom_hline(yintercept = produpper, lty= 2, colour="darkgrey") + 
    geom_hline(yintercept = prodlower, lty= 2, colour="darkgrey") + 
    geom_vline(xintercept = hi, lty=1, colour="black") +
    geom_vline(xintercept = lo, lty=1, colour="black") +
    geom_point(dat = example_calibration_output$output_c0$data, aes(x=Prey, y = ChicksPerNest), colour="red") +
    geom_line(dat = example_calibration_output$output_c0$data, aes(x=Prey, y = ChicksPerNest), colour="red") +
    geom_smooth(dat = example_calibration_output$output_c0$data, aes(x=Prey, y = ChicksPerNest), colour="orange",method = "loess") +
    labs(y = "Chicks Per Nest")
  
  pBoth <- grid.arrange(pAd, pCh, ncol=1)




```
