---
title: "Example seabORD run"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Example seabORD run}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

Load the seabORD package

```{r setup, message = FALSE}

#load seabORD
library(seabORD)

library(magrittr)

```

Before following this example, the model is assumed to be already calibrated. If not follow the article on calibration before running this example. hhhghgh

```{r, echo=FALSE}

### UK outline

uk_map <- dplyr::select(
  rnaturalearth::ne_countries(country = "United Kingdom",
                              scale = "medium",
                              returnclass = "sf"),
  name,
  geometry)


### theme definition
sysfonts::font_add_google("Montserrat", "montserrat")  # Base and heading font
sysfonts::font_add_google("JetBrains Mono", "jetbrains_mono")  # Code font
showtext::showtext_auto()

theme_bslib <- ggplot2::theme(
  plot.background = ggplot2::element_rect(fill = "#fff", color = NA),       # Background color
  panel.background = ggplot2::element_rect(fill = "#EAEFEC", color = NA),  # Panel background (secondary)
  panel.grid.major = ggplot2::element_line(color = "#EAEFEC"),             # Major grid lines
  panel.grid.minor = ggplot2::element_line(color = "#EAEFEC"),             # Minor grid lines
  axis.text = ggplot2::element_text(color = "#292C2F", family = "montserrat"),  # Axis text (foreground and font)
  axis.title = ggplot2::element_text(color = "#292C2F", family = "montserrat"), # Axis titles
  plot.title = ggplot2::element_text(color = "#0483A4", family = "montserrat", size = 16, face = "bold", hjust = 0.5), # Title
  legend.background = ggplot2::element_rect(fill = "#fff", color = NA),    # Legend background
  legend.key = ggplot2::element_rect(fill = "#EAEFEC", color = NA),        # Legend key
  legend.text = ggplot2::element_text(color = "#292C2F", family = "montserrat"), # Legend text
  plot.caption = ggplot2::element_text(color = "#292C2F", family = "jetbrains_mono"),
  axis.title.x = ggplot2::element_blank(),  # Remove x-axis label
  axis.title.y = ggplot2::element_blank(),# Caption for code font
)
```

# Preparing the inputs

To run seabORD several inputs are needed:

| arguments | type | description |
|----|----|----|
| Par | List | The main parameters controlling/defining this run |
| modPar | List | Parameters relating to the model mode and computer environment |
| ordPar | List | Input parameters relating to the ORDs |
| switches | List | A set of switches/flags used to control optional features of the run |
| seamask | Raster | land/sea grid, 1km expected. |
| spadat1 |  | description TBC |
| spadat2 |  | description TBC |
| spdat |  | Species-specific parameters |
| BrdData |  | description TBC |
| FrgCompData |  | description TBC |
| fltdist_base |  | Flight distance by sea, without ORDs. See user guide for details. |
| FlightGridcorrection |  | Flight distance transition layer (gdistance) |
| ORDpoly |  | The ORD footprints |

The package comes with some example input data to run the model with. Here's an example run of the model with detailed explanation for each input argument.

## Par

List for biological parameters:

-   thisSpecies:

-   colonies:

-   Nscalefactor:

-   Prob_Displacement

-   Prob_Barrier

-   PreyType

-   collision

-   SiteSelectionMethod

-   MaxDistancekm

-   PropInRange

-   Npairspercol

-   Pmedian

```{r}
Par_example <-  seabORD::example_1_lists$Par

str(Par_example)
```

## modPar

List for model parameters.

-   Nparallel

-   initialseed

-   reference

-   outputdir

-   Nreplicates

make a folder in the directory to store your output (you can place it wherever)

```{r}
# Create the ouput folder "output_seabORD" if not existing already
if (!dir.exists("output_seabORD")) {
  dir.create("output_seabORD")
}
```

and then the list of modPar

```{r}
modPar_example <- seabORD::example_1_lists$modPar

str(modPar_example)
```

## ordPar

List for wind farm parameters

-   include_ORDs

-   parnames

-   FootprintBorder

-   BufferZone

```{r}
ordPar_example <-  seabORD::example_1_lists$ordPar

str(ordPar_example)
```

## Switches

This parameter requires a llist() object that defines the optional features of the run. It accepts the following options:

-   **environment** : "serial" or "parallel"

-   **modelmode**: "scenario" or "calibration"

-   **debugmode**: 0, 1, or 2

-   **bycol**: TRUE or FALSE

-   **bysus**: TRUE or FALSE

-   **bych**: TRUE or FALSE

-   **printdaily**: TRUE or FALSE

-   **printseason**: TRUE or FALSE

-   **printpair**: TRUE or FALSE

-   **minout**: TRUE or FALSE

-   **silent**: TRUE or FALSE

-   **saverds**: TRUE or FALSE

-   **savebirdflightmap**: TRUE or FALSE

There is an example list available in the package that can be accessed by simply call in the built-in list "example_switch_list":

```{r}

#get the example switches list
switches_example <- seabORD::example_1_lists$switches

#check the setting of this example
str(switches_example)
```

In this example seabORD will in serial and in scenario mode; without debugging and by saving the output as a .rds file.

## seamask

Example data to be used as seamask can be found built in the package as "seamask_3035_example" to make the raster for the seamask:

```{r}
example_data_seamask <- seabORD::seamask_3035_example
str(example_data_seamask)
```

We can use this data to make the raster using the terra package

```{r, warning=FALSE, message=FALSE}

#make a raster with the right dimension
seamask_example <- terra::rast(
  nrows = example_data_seamask$metadata[["n_rows"]],
  ncols = example_data_seamask$metadata[["n_cols"]],
  xmin = example_data_seamask$metadata[["x_min"]],
  xmax = example_data_seamask$metadata[["x_max"]],
  ymin = example_data_seamask$metadata[["y_min"]],
  ymax = example_data_seamask$metadata[["y_max"]],
  crs = example_data_seamask$metadata[["crs"]]
)
#fill it with the data
terra::values(seamask_example) <- example_data_seamask$matrix

```

which looks like

```{r, echo = FALSE}

seamask_example_repr <- terra::project(seamask_example, 
                                       terra::crs(uk_map))


seamask_df <- as.data.frame(seamask_example_repr, xy = TRUE)


# Plot
ggplot2::ggplot() +
  ggplot2::geom_raster(data = seamask_df, ggplot2::aes(x = x, y = y, fill = lyr.1)) +
  #uk outline
 ggplot2::geom_sf(data = uk_map, fill = NA, color = "#D63333", size = 0.5) +
ggplot2::scale_fill_gradient(
    low = "#0483A4",   # Color for 0
    high = "#0483A4",  # Keep the same for continuous scale since values are only 0
    na.value = "gray80",  # Color for NA values
    guide = ggplot2::guide_colorbar(na.translate = TRUE, title = "Value")  # Add NA to legend
  ) +
 theme_bslib +  #palette theme
 ggplot2::labs(fill = "Seamask Raster Value",
               title = "Raster and outline of the UK (red)",
               caption = "Example 1: data seamask")
```

make it a raster (not sustainable)

```{r}
seamask_example <- raster::raster(seamask_example)
names(seamask_example) <- "seamask_3035" 
```

## spadat1

For this example the values for spadat1 for the site "UK9004171" are taken from the built-in dataset "spacoordinates" which has data for all the sites relevant for the CEF. Get only what's relevant for this example.

```{r}
#open the dataset provided with the pacakge
spadat1_example <- 
  tibble::as_tibble( #as a tibble
      dplyr::filter(seabORD::spacoordinates, #the full dataset
                    SITECODE == "UK9004171") #the iste of interest
  )

print(t(spadat1_example)) #display the data for the example site
```

## spadat2

Can be derived from "spalist"

```{r}
spadat2_example <- 
  tibble::as_tibble( #as a tibble
    dplyr::filter(seabORD::spalist, #the whole list
                  SITE_CODE == "UK9004171") #the site of interest
    )

#take a look
print(t(spadat2_example))
```

## spdat

The species specific data for the Black-legged Kittiwake be derived from the built in dataset "energeticsandpreydata" that has data for all 4 species that seabORD models.

```{r}
spdat_example <- 
   #tibble::as_tibble( #as a tibble
     dplyr::filter(seabORD::energeticsandpreydata,
                   Code == "KI")
    # )
    
#take a look
print(t(spdat_example))
```

## BrdData

there is a built in example dataset for this run. This is a raster file that must re-constructed from the data as follows

```{r}
example_data_brd <- seabORD::BrdData_example
#make the raster with the right dimensions..
BrdData_example <- terra::rast(
  nrows = example_data_brd$metadata[["n_rows"]],
  ncols = example_data_brd$metadata[["n_cols"]],
  xmin = example_data_brd$metadata[["x_min"]],
  xmax = example_data_brd$metadata[["x_max"]],
  ymin = example_data_brd$metadata[["y_min"]],
  ymax = example_data_brd$metadata[["y_max"]],
  crs = example_data_brd$metadata[["crs"]]
)
#fill it with the data
terra::values(BrdData_example) <- example_data_brd$matrix
```

Which has data that plot like this:

```{r, echo=FALSE}

BrdData_example_repr <- terra::project(BrdData_example, 
                                       terra::crs(uk_map))
BrdData_df <- as.data.frame(BrdData_example_repr, xy = TRUE)


ggplot2::ggplot() +
  # Plot the raster
  ggplot2:: geom_raster(data = BrdData_df, ggplot2::aes(x = x, y = y, fill = ifelse(lyr.1 == 0, NA, lyr.1))) +
  # Add the reprojected UK boundary
  ggplot2::geom_sf(data = uk_map, fill = NA, color = "#D63333", size = 1.5) +
  # Customize the plot
  theme_bslib +  #palette theme
    ggplot2::scale_fill_viridis_c(
    na.value = "gray",  # Set the color for NA (i.e., for 0 values)
    name = "BrdData Data", direction = -1) +
  ggplot2::labs(fill = "BrdData Raster Value",
               title = "Raster and outline of the UK (red)",
               caption = "Example 1: data BrdData")
```

as a raster

```{r}
BrdData_example <- raster::raster(BrdData_example)
names(BrdData_example) <- "Forth.Islands"
```

## FrgCompData

Also example data available

```{r}
example_data_frg <- seabORD::frgcompdata_example
#make the raster with the right dimensions..
FrgCompData_example <- terra::rast(
  nrows = example_data_frg$metadata[["n_rows"]],
  ncols = example_data_frg$metadata[["n_cols"]],
  xmin = example_data_frg$metadata[["x_min"]],
  xmax = example_data_frg$metadata[["x_max"]],
  ymin = example_data_frg$metadata[["y_min"]],
  ymax = example_data_frg$metadata[["y_max"]],
  crs = example_data_frg$metadata[["crs"]]
)
#fill it with the data
terra::values(FrgCompData_example) <- example_data_frg$matrix
```

which looks like:

```{r, echo=FALSE}
FrgCompData_example_repr <- terra::project(FrgCompData_example, 
                                           terra::crs(uk_map))
FrgCompData_df <- as.data.frame(FrgCompData_example_repr, xy = TRUE)

ggplot2::ggplot() +
  # Plot the raster
  ggplot2:: geom_raster(data = FrgCompData_df, ggplot2::aes(x = x, y = y, fill = ifelse(lyr.1 == 0, NA, lyr.1))) +
  # Add the reprojected UK boundary
  ggplot2::geom_sf(data = uk_map, fill = NA, color = "#D63333", size = 1.5) +
  # Customize the plot
  theme_bslib +  #palette theme
    ggplot2::scale_fill_viridis_c(
    na.value = "gray",  # Set the color for NA (i.e., for 0 values)
    name = "FrgCompData Data", direction = -1) +
  ggplot2::labs(fill = "FrgCompData Raster Value",
               title = "Raster and outline of the UK (red)",
               caption = "Example 1: data FrgCompData")

#plot(FrgCompData_example)
```

as raster

```{r}
FrgCompData_example <- raster::raster(FrgCompData_example)
names(FrgCompData_example) <- "Forth.Islands"
```

## fltdist_base

data available for the example; we construct it as a raster

```{r}
UK9004171_bysea <- seabORD::UK9004171_bysea_3035
#make the raster with the right dimensions..
fltdist_base_example <- terra::rast(
  nrows = UK9004171_bysea$metadata[["n_rows"]],
  ncols = UK9004171_bysea$metadata[["n_cols"]],
  xmin = UK9004171_bysea$metadata[["x_min"]],
  xmax = UK9004171_bysea$metadata[["x_max"]],
  ymin = UK9004171_bysea$metadata[["y_min"]],
  ymax = UK9004171_bysea$metadata[["y_max"]],
  crs = UK9004171_bysea$metadata[["crs"]]
)
#fill it with the data
terra::values(fltdist_base_example) <- UK9004171_bysea$matrix
```

which looks like:

```{r, echo = FALSE}

fltdist_base_example_repr <- terra::project(fltdist_base_example, 
                                           terra::crs(uk_map))
fltdist_base_df <- as.data.frame(fltdist_base_example_repr, xy = TRUE)

ggplot2::ggplot() +
  # Plot the raster
  ggplot2:: geom_raster(data = fltdist_base_df, ggplot2::aes(x = x, y = y, fill = ifelse(lyr.1 == 0, NA, lyr.1))) +
  # Add the reprojected UK boundary
  ggplot2::geom_sf(data = uk_map, fill = NA, color = "#D63333", size = 1.5) +
  # Customize the plot
  theme_bslib +  #palette theme
    ggplot2::scale_fill_viridis_c(
    na.value = "gray",  # Set the color for NA (i.e., for 0 values)
    name = "fltdist_base Data", direction = -1) +
  ggplot2::labs(fill = "fltdist_base Raster Value",
               title = "Raster and outline of the UK (red)",
               caption = "Example 1: data fltdist_base")
```

as a ratser brick

```{r}
fltdist_base_example <- raster::brick(fltdist_base_example)
names(fltdist_base_example) <- "UK9004171"
```

## FlightGridcorrection

It's a TransitionLayer class object. One example provided that can be loaded

```{r}
FlightGridcorrection <- seabORD::FlightGridcorrection_3035

```

and looks like:

```{r, echo=FALSE}
# to plot it you have to extract the values back into a raster):

# get the Transition Matrix
transition_matrix <- gdistance::transitionMatrix(FlightGridcorrection)

#non_zero_count <- Matrix::nnzero(transition_matrix)

# extract the diagonal values
#diag_vals <- Matrix::diag(transition_matrix)

#Sum each row to calculate total connectivity for each cell
total_connectivity <- Matrix::rowSums(transition_matrix)

#Create a Raster for the total_connectivity
total_connectivity_raster <- terra::rast(
  nrows = nrow(FlightGridcorrection),
  ncols = ncol(FlightGridcorrection),
  xmin = raster::extent(FlightGridcorrection)[1],
  xmax = raster::extent(FlightGridcorrection)[2],
  ymin = raster::extent(FlightGridcorrection)[3],
  ymax = raster::extent(FlightGridcorrection)[4],
  crs = "+proj=laea +lat_0=52 +lon_0=10 +x_0=4321000 +y_0=3210000 +ellps=GRS80 +units=m +no_defs",
  vals = total_connectivity #assign values
)

#change crs
total_connectivity_raster <- terra::project(total_connectivity_raster,
                                            terra::crs(uk_map))  

total_connectivity_df <- as.data.frame(total_connectivity_raster, xy = TRUE)

ggplot2::ggplot() +
  # Plot the raster
  ggplot2:: geom_raster(data = total_connectivity_df, ggplot2::aes(x = x, y = y, fill = ifelse(lyr.1 == 0, NA, lyr.1))) +
  # Add the reprojected UK boundary
  ggplot2::geom_sf(data = uk_map, fill = NA, color = "#D63333", size = 1.5) +
  # Customize the plot
  theme_bslib +  #palette theme
    ggplot2::scale_fill_viridis_c(
    na.value = "gray",  # Set the color for NA (i.e., for 0 values)
    name = "total_connectivity_df Data", direction = -1) +
  ggplot2::labs(fill = "total_connectivity_df Raster Value",
               title = "Raster and outline of the UK (red)",
               caption = "Example 1: data total_connectivity_df")
```

## ORDpoly

```{r}
ORDpoly_example <- seabORD::ORDpoly_example
```

whit this multipolygon shapes:

```{r, echo=FALSE}
ORDpoly_example_repr <- ORDpoly_example %>%
  sf::st_transform(sf::st_crs(uk_map))

ggplot2::ggplot() +
  # Plot the reprojected UK map in red
  ggplot2::geom_sf(data =  uk_map, fill = NA, color = "#D63333", size = 1.5) +
  # Add the ORDpoly_example shapes with a different color
  ggplot2::geom_sf(data = ORDpoly_example_repr, fill = "#0483A4", color = "#0483A4", size = 1) +
  ggplot2::coord_sf(xlim = c(-4, 1), ylim = c(55, 57)) +
  theme_bslib +  #palette theme
  ggplot2::labs(title = "UK Map (red) and ORDpoly Shapes (blue)")

```

# Running seabORD

Now that all the inputs are ready we can run the model using:

```{r}

# SeabORDoutput <- seabord(Par = Par_example,
#                          ordPar = ordPar_example,
#                          modPar = modPar_example,
#                          switches = switches_example,
#                          seamask = seamask_example,
#                          spadat1 = spadat1_example,
#                          spadat2 = spadat2_example,
#                          spdat = spdat_example,
#                          BrdData = BrdData_example,
#                          FrgCompData = FrgCompData_example,
#                          fltdist_base = fltdist_base_example,
#                          FlightGridcorrection = FlightGridcorrection,
#                          ORDpoly = ORDpoly_example)
```

the run can take some time..\
the output will be saved in in your output directory set up in modPar.

To learn more about the main functions behind seabORD you can read the article on the main functions

To see how you can read in and plot the results you can read the article on how to interpret the results.
